---
title: "Multivariate Analysis Project"
author: "Nishanth Girish"
date: "2024-01-02"
output: word_document
---


```{r}
# Data
library(httr)
data.full <-read.csv("C:/Users/nisha/OneDrive/Documents/Multivariate/skulls.csv")

#data.full <- data.full[,-1]

data <- data.full[,2:5]
data
```



```{r}
# Mean vectors;

# data stat summary #
x_bar<- apply(data, 2, mean)  
x_bar
```


```{r}
# Covariance matrix
cov.mat <- cov(data)
cov.mat
```


```{r}
skulls1 <- as.matrix(subset(data.full, data.full[,1]=="c4000BC")[,2:5])
skulls2 <- as.matrix(subset(data.full, data.full[,1]=="c3300BC")[,2:5])
skulls3 <- as.matrix(subset(data.full, data.full[,1]=="c1850BC")[,2:5])
skulls4 <- as.matrix(subset(data.full, data.full[,1]=="c200BC")[,2:5])
skulls5 <- as.matrix(subset(data.full, data.full[,1]=="cAD150")[,2:5])

# group mean #
x_bar_1 <- apply(skulls1, 2, mean)
x_bar_2 <- apply(skulls2, 2, mean)
x_bar_3 <- apply(skulls3, 2, mean)
x_bar_4 <- apply(skulls4, 2, mean)
x_bar_5 <- apply(skulls5, 2, mean)

# Mean Vectors ( mean of each variable for each time period)
cbind(x_bar_1, x_bar_2, x_bar_3, x_bar_4) 
```
```{r}
# Correlation matrix

corr.mat <- cor(data)
corr.mat                          # None of the variables are highly correlated with each other.
```

```{r}
# correlation
attach(data)
library(corrplot)
corrplot(cor(data),method = 'number')
```

```{r}
# group correlations #
corr1 <- cor(skulls1)
corr2 <- cor(skulls2)
corr3 <- cor(skulls3)
corr4 <- cor(skulls4)
corr5 <- cor(skulls5)

# group covariances #
s1 <- cov(skulls1)
s2 <- cov(skulls2)
s3 <- cov(skulls3)
s4 <- cov(skulls4)
s5 <- cov(skulls5)
```


```{r}
# Covariance Matrices
s1
s2
s3
s4
s5
```

```{r}
# Correlation matrices
corr1
corr2
corr3
corr4
corr5
```

# Now we will be checking for the assumptions of MANOVA:
```{r}
# 1. Assumption of Homogeneity of Covariance Matrices

# Box's M test for equality of covariance matrices.
library(biotools)
boxM(data, data.full$epoch)
# Fail to reject null hypothesis. Conclude we have equal covariance matrices
```
-> We fail to reject the null hypothesis of homogeneity of covariance matrices.


```{r}
# 2. (i)Assumption of normality
# Checking for the mormality of each of the variables individually (univariate normality)
par(mfrow =c(2,2))
hist(data$mb)
hist(data$bh)
hist(data$bl)
hist(data$nh)

shapiro.test(data$mb)
shapiro.test(data$bh)                    
shapiro.test(data$bl)
shapiro.test(data$nh)
# p-values > 0,05

# Set up a 2x2 grid for subplots
par(mfrow = c(2, 2))
qqnorm(data$mb, main="mb")
qqline(data$mb)

qqnorm(data$bh, main="bh")
qqline(data$bh)

qqnorm(data$bl, main ="bl")
qqline(data$bl)

qqnorm(data$nh, main = "nh")
qqline(data$nh)
```
Approximately normal.
-> We fail to reject the null hypothesis that observations come from a normal distribution.

```{r}
# (ii) Checking for normality assumption: (multivariate normality)

# assessment of normality #
library(nortest)
nor.test <- matrix(ncol=1,nrow=ncol(data))
for ( i in 1:ncol(data)){
  nor.test[i] <- ad.test(data[,i])$p.value        #Anderson-Darling test
}
nor.test

# Normality test for multivariate variables #
library(mvnormtest)

#H0: Normal
#H1: Not normal
mshapiro.test(t(data))  # p-value > 0.05. Fail to reject H0. -> Normal
```

-> The data follows multivariate normal distribution.

```{r}
# scatterplots #
# option 1 #
library(lattice)
splom(data.full[,2:5], groups=data.full[,1],pch=c(1,2,3,4,5), col=c(1,2,3,4,5), key=list(title="Five Time Periods",columns=5,points=list(pch=c(1,2,3,4,5),col=c(1,2,3,4,5)),text=list(c("4000 B.C.","3300 B.C.","1850 B.C.", "200 B.C", "150 A.D"))))

# option 2 #
library(car)
scatterplotMatrix(~mb+bh+bl+nh|epoch, data=data.full, col=c(1,2,3,4,5), pch=c(1,2,3,4,5),main="Five Time Periods")

```
Approximately normal

```{r}
# Assumption of no multi-collinearity

cor(data)      # None of the variables are highly correlated with each other


library(mctest)
mlr1 = lm(mb ~ ., data = data)
mlr2 = lm(bh ~ ., data = data)
mlr3 = lm(bl ~ ., data = data)
mlr4 = lm(nh ~ ., data = data)

# VIF values:
imcdiag(mlr1)   # Individual Multicollinearity Diagnostics
imcdiag(mlr2)
imcdiag(mlr3)
imcdiag(mlr4)
```
-> We observe a low VIF value. Hence, there exists no multicollinearity.

```{r}
#Outliers 

# Multivariate Outlier Detection #
library(mvoutlier)
aq.plot(data, delta=qchisq(0.975, df=ncol(data)), quan=1/2, alpha=0.05)
```
Outliers based on 97.5% quantile: observations #4, 34, 78, 131, 149


```{r}
# Outlier detection using Mahalanobis distance:

# Specify the mean and covariance matrix
mean_vec <- colMeans(data)
cov_matrix <- cov(data)

# Calculate Mahalanobis distance
D2 = mahalanobis_dist <- mahalanobis(data, center = mean_vec, cov = cov_matrix)
print(D2)
# Plot
plot(density(D2, bw = 0.5),
     main="Squared Mahalanobis Distances") ; rug(D2)

# QQ plot
qqplot(qchisq(ppoints(100), df = 4), D2,
       main = expression("Q-Q plot of Mahalanobis" * ~D^2 *
                         " vs. quantiles of" * ~ chi[4]^2))
abline(0, 1, col = 'gray')

# Outliers:
# Identify observations with D2 greater than 9.7
outliers <- which(D2 > 9.7)
outliers

# Based on the Mahalanobis distance, observations 4, 29  34  78 123 131 146 147 149 are outliers.
```


```{r}
# Box Plots
attach(data.full)

# Reorder the levels of the Group variable
data.full$epoch <- factor(data.full$epoch, levels = c("c4000BC", "c3300BC", "c1850BC", "c200BC","cAD150"))

par(mfrow = c(2, 2))
boxplot(data.full$mb ~ data.full$epoch, main = "Maximum Breadth of Skull", xlab = "Time Period")
boxplot(nh ~ epoch, main = "Nasal Height of Skull",xlab = "Time Period")
boxplot(bl ~ epoch, main = "Basialveolar Length of Skull", xlab = "Time Period")
boxplot(bh ~ epoch, main = "Basibregmatic Height of Skull", xlab = "Time Period")
```
From the Box plots, it is evident that the different dimensions of the skulls differ over the different time periods.


```{r}
# MANOVA #
fit.lm <- lm(cbind(mb, bh, bl, nh)~epoch , data = data.full)
fit.manova <- Manova(fit.lm)
summary(fit.manova)
```
# p-value < 0.05. (Statistically significant). Group means differ significantly. Skull dimensions change over the time periods.

```{r}

manova1 = manova(cbind(mb, bh, bl, nh)~epoch , data = data.full)
#summary(manova1)

summary.aov(manova1)  #for each of the response variable
```
-> For the variables mb, bh, bl the p-value is less than 0.05. Hence, we reject the null hypothesis that the means of the groups does not differ significantly over the time periods.
-> For the variable nh, the p-value is high, hence we fail to reject to the null hypothesis that the mean nasal height does not differ significantly over the time periods.

###############################################################################################
# k-means clustering
```{r}
# Elbow method
wss <- numeric(10)
for (i in 1:10) {
  kmeans_model <- kmeans(scaled_data, centers = i)
  wss[i] <- kmeans_model$tot.withinss
}
plot(1:10, wss, type = "b", xlab = "Number of Clusters (k)", ylab = "Within-cluster Sum of Squares")

```
Optimum number of clusters is 5. 

```{r}
# Loading package 
library(ClusterR) 
library(cluster) 

# scaling the data
scaled_data <- scale(data)

# Fitting K-Means clustering Model  
set.seed(240) # Setting seed 
kmeans.re <- kmeans(scaled_data, centers = 5) 
kmeans.re 

# Cluster identification for each observation 
kmeans.re$cluster 

# Confusion Matrix 
cm <- table(data.full$epoch, kmeans.re$cluster) 
cm

# Silhouette score:
silhouette_plot <- silhouette(kmeans.re$cluster, dist(scaled_data))
plot(silhouette_plot, main = "Silhouette Plot for K-Means Clustering")

# Example scatter plot for the first two variables
plot(scaled_data[, 1], scaled_data[, 2], col = kmeans.re$cluster , pch = 16, main = "K-Means Clustering Results")
points(kmeans.re$centers[, 1], kmeans.re$centers[, 2], col = 1:5, pch = 8, cex = 2)  # Adds cluster centers


```
```{r}
# Cluster plot
clusplot(data[,c("mb","bh","bl","nh")],
          kmeans.re$cluster,
         lines = 0,
         shade = TRUE, 
         color = TRUE,
         labels = 1,
         plotchar = TRUE, 
         span = TRUE, 
         main = paste("Clusters"))
         
```



